---
title: "Extending ggplot2 for linked and animated web graphics"
documentclass: article
fontsize: 12pt
geometry: margin=1in
linkcolor: cyan
citecolor: black
bibliography: refs.bib
abstract: Interactive web graphics are great for communication and knowledge sharing, but are difficult to leverage during the exploratory phase of a data science workflow. Even before the web, interactive graphics helped data analysts quickly gather insight from data, discover the unexpected, and develop better model diagnostics. Web technologies, however, are not designed to fit inside an exploratory data analysis workflow where rapid iteration between data manipulation, modeling, and visualization must occur. We propose the R package **animint** for rapid creation of linked and animated web graphics through a simple extension of **ggplot2**'s implementation of the Grammar of Graphics. The extension allows users to use their existing **ggplot2** code, then leverage an idomatic API for describing animations and graphical queries between multiple linked views in a standalone web page.
keywords: Animation, Multiple linked views, Statistical graphics, Exploratory data analysis, Web technologies
output:
  bookdown::pdf_book:
    template: jcgs-template.tex
    toc: no
    keep_tex: yes
    citation_package: none
---


# Introduction {#intro}

<!--
%% In general, there are three influential roles involving an Animint
%% visualization: the developer, who implements the Animint library; the
%% designer, who uses the Animint library to define a visualization; and
%% the user, who selects data subsets to view in a web browser. The main
%% goal of Animint is to provide an expressive language for designers,
%% while allowing users the freedom to interact with the plot to
%% selectively view data subsets of interest. The designer specifies data
%% sets and maps variables to interactive visual elements using the
%% Animint DSL, then uses the Animint library to compile and save an
%% interactive animation. The user writes no code, but can view and
%% interact with an Animint visualization by clicking Scalable Vector
%% Graphics (SVG) elements in a web browser. The Animint library
%% developer is responsible for the plot rendering details, which allows
%% the others to focus on designing and consuming visualizations.

<<opts, echo=FALSE, message=FALSE, cache=FALSE>>=
knitr::opts_chunk$set(tidy = FALSE, eval = FALSE)
knitr::opts_knit$set(out.format = "latex")
knitr::knit_theme$set("edit-matlab")
@


%%Interactive and dynamic statistical graphics toolkits play an important role in the exploratory phase of a data analysis cycle. The web is an attractive platform for interactive graphics thanks to its portability and accessibility, but due to a lack of tools for quick iteration, web graphics are not a practical tool for exploring data. Instead, they are mostly used for the presentation of findings (i.e., when the 'correct' visualization is already known). This is a source of conflict, as one often needs different tools depending on whether the work is exploratory or intended for public consumption, which results in wasted effort.

%%The R package \texttt{ggplot2} is a popular tool for data visualization due in part to its elegant design inspired by the grammar of graphics. The fundamental concept in the grammar of graphics is that any graphic can be described through a set of mappings from data variables to visual space, referred to commonly as \texttt{aesthetic} mappings. The grammar of graphics and ggplot2 were designed for static graphics
-->

For more than a half century now, statisticians have designed, built, and used 
interactive graphics for exploring high-dimensional data and better informing
their modeling process.
In fact, the ASA maintains a video library (<http://stat-graphics.org/movies/>) 
to document and demonstrate applications of instrumental interactive 
statistical graphics systems such as  `PRIM-9` [@PRIM9], `Data Viewer` 
[@viewing-pipeline], `XGobi` [@xgobi], `GGobi` [@ggobi:2007], and 
`Mondrian` [@mondrian]. These, as well as other influential systems, such as 
`LISP-STAT` [@LISP-STAT] and `MANET` [@MANET], all have a rich support for 
accomplishing a wide variety of statistical analysis tasks, and most were 
developed before the web browser had rich graphics support.

All of these systems, as well as some more modern systems, such as
**rggobi** [@rggobi], **iplots** [@Urbanek2011], **loon** [@loon], etc, require 
a heavy set of computational dependencies in order to view or interact with 
graphics. These requirements grant the freedom to leverage libraries with 
sophisticated statistical functionality on-demand, but it limits the 
ability to share or embed such graphics in a larger document. Some of these
systems allow users to create the graphics from the command-line, which 
as @Unwin:1999vp points out, allows
power users to combine the strengths of a programming interface 
(e.g., precise, repeatable, fast, and extensible) with the 
strengths of a graphical interface (e.g., intuitive, forgiving, and easy-to-use). 
Web technologies can certainly be used to build a similar class 
of system, but in order to capitalize on the key strength of web technologies 
(e.g., accessible, portable, and composable), we must be mindful if which
technologies we're requiring in such a system, and minimize thosee requirements 
whenever possible.

Generally speaking, web graphics that use purely client-side
technologies (i.e., `HTML`, `SVG`, `CSS`, and `JavaScript`) are desired over 
client-server web applications because of their relative ease of distribution
and maintenance. This is why many web-based graphing libraries work entirely 
with client-side technologies, like Vega [@vega] and plotly.js [@plotly].  
Unfortunately, client-side technologies are not particularly well-suited for 
statistical computation, which we often want to leverage via dynamic controls 
in an interactive statistical graphics system. In this scenario, it often makes
sense to introduce a client-server infrastructure to leverage functionality that 
is not natively supported by web browsers (e.g., `R`, `python`, etc).

<!--
Table X gives a high-level depiction of the different infrastructure required 
for a purely client-side (i.e. standalone) webpage versus a client-server web 
application. 
-->

Focusing just on the R language, there are now numerous ways
to develop web applications, including the R package **shiny** [@shiny], 
which makes it easy for R users to take their existing scripting workflow 
and wrap a web interface around it. 
**Shiny** is great for quickly prototyping interactive webpages
that re-execute R code on-demand, but that flexibility comes at the cost
of requiring a complex web server framework, which can be hard to scale, 
maintain, and secure sensitive information. Unfortunately, all too often,
a web application framework is used to implement linked and animated graphics 
that could more easily be described with an idiomatic R interface which produces 
a purely client-side result. 

There are now a large number of R packages that interface with purely client-side
graphing libraries and give users the option of embedding these graphics in a 
larger web application. In fact, this is a large enough use case
that the R package **htmlwidgets** [@htmlwidgets] was created to make it easier 
to get these interfaces to work seamlessly in any context 
(e.g., **shiny**, **rmarkdown**, RStudio, R prompt, Jupyter notebook, etc). 
In our experience, these R packages rarely provide a way to describe linked 
views in R and produce a purely client-side result, which we view as a powerful 
workflow for "production-ready" exploratory web graphics. The R package 
**crosstalk** is a nice step in this direction

<!-- 
TODO: could try to make the argument that projects like Vega-Lite, that try to
define a JSON specification for linked view are too seperated from the semantics 
of the data. 
-->

We propose an extension of **ggplot2**'s layered Grammar of Graphics API to 
create interactive web graphics that don't require server-side technologies. The 
core idea lies in attaching metadata to graphical marks that can be used to 
hide/show subsets of data. The resulting framework is quite similar to what @Cook:2007uk
describes as brushing in multiple linked views as a database query. The assignment
of data to graphical marks is done through *aesthetic mappings*, which is a term
the Grammar of Graphics uses for mapping data to visual attributes (e.g., color, 
shape, x, y, etc). Typically aesthetic mappings are visual, meaning they can be
easily seen in a static graphic, but our proposed variable mappings are more 
conceptual, meaning that they can't easily be seen, but can be used to control 
certain interactive properties. To give a small example, Figure \@ref(fig:tips) 
depicts a graphical query made by assigning sensible metadata to graphical marks 
via the `clickSelects` and `showSelected` aesthetics.

In addition to graphical queries, our extension supports a number of other
interactive features, including animation, tooltips, and hyperlinks.
A summary of these extensions and relevant additional options are provided 
in Table \@ref(tab:overview). There are a number of other options that can 
be used to control things specific to our implementation in the R package
**animint** which is described in the supplemental materials.

\begin{table}

\caption{
New features that animint adds to the grammar of graphics.
}\label{tab:overview}
\small
\begin{tabularx}{\textwidth}{|l|l|l|}
\hline
Feature & Type & Description \\
\hline
\texttt{clickSelects} & aesthetic & value(s) to select on click \\
\texttt{showSelected} & aesthetic & value(s) attached to mark(s) that determine when they are shown  \\
\texttt{tooltip} & aesthetic & information to display on hover \\
\texttt{href} & aesthetic & URL link to open on click \\
\texttt{selector.types} & option & should selections accumulate?  \\
\texttt{first} & option & what value(s) should be selected by default?  \\
\texttt{time} & option & delay between animation frames.  \\
\texttt{duration} & option & to specify smooth transitions.  \\
\texttt{key} & aesthetic & value(s) attached to mark(s) for smooth transitions. \\
\texttt{selectize} & option & include a dropdown widget to set selection value(s) indirectly? \\
\hline
\end{tabularx}

\end{table}




<!--
Interactive graphics are an important part of applied statistical
practice, because they can lead to a better understanding of
high-dimensional data sets and models. Interactive graphics are useful
in the context of EDA, model selection/validation, and presenting
results of an analysis. Interactive graphics toolkits in R have been
available for decades, but these approaches are often not easy to
reproduce or distribute to a larger audience. It is true that most
graphics generated during EDA are ultimately not useful, but
sometimes, understanding gained during this phase is most easily
shared via the interactive graphics themselves. Thus, there is value
in being able to easily share and embed interactive graphics inside a
larger report. Unfortunately, this is typically hard, if not
impossible, using traditional interactive graphics toolkits. As a
result, there is a large disconnect between the visualization tools
that we use for exploration versus presentation.

One of the most widely used R packages is **ggplot2** [@ggplot2-book], a data
visualization package inspired by the grammar of graphics [@wilkinson]. 
In fact, @Donoho:2015tu writes: "This effort may have more impact on 
today’s practice of data analysis than many highly-regarded theoretical 
statistics papers". In our experience, **ggplot2** has made an impact thanks to its
foundation in the grammar of graphics, carefully chosen defaults, and overall
usability. This helps data analysts rapidly iterate and discover informative
visualizations -- an essential task in exploratory data analysis (EDA). When 
dealing with high-dimensional data, however, it is often useful to produce 
interactive and/or dynamic graphics, which **ggplot2** does not inherently support. 

We aim to narrow this gap in visualization tools by extending **ggplot2**'s grammar 
of graphics implementation for interactive and dynamic web graphics. Our 
extension allows one to create animated transitions
and perform dynamic queries via direct manipulation of linked views like those
described in @Ahlberg:1991 and @Buja:1991vh. 
A conceptual model for our extension is provided in
Section \@ref(extension) and Section \@ref(animation). In 
Section \@ref(worldbank), we demonstrate our extension with an example. 
In Section \@ref(implementation), we outline design decisions made in our
implementation in the R package **animint**. In Section \@ref(performance), we
provide a sense of the scope of our system and its performance limitations through a
handful of examples. In Section \@ref(compare), we conduct a comparison
study by replicating examples with other leading systems. Finally, in 
Section \@ref(limitations), we discuss future work and limitations of
our current system.

-->

# Related work

<!--
We aim to provide a system which empowers **ggplot2** users to go beyond the 
confines of static graphics with minimal friction imposed upon their current
workflow. We acknowledge that numerous systems which support similar 
visualization techniques exist outside of the R ecosystem, but we 
intentionally focus on R interfaces since the surrounding statistical 
computing environment is crucial for enabling an efficient EDA workflow.
-->

It is important to acknowledge that **ggplot2** is built on top of the R package 
**grid**, a low-level graphics system, which is now bundled with R itself 
[@RCore]. Neither **grid**, nor **base** R graphics, have strong support for 
handling user interaction, which creates a need for add-on packages. There are a 
number of approaches these packages take to rendering, each with their own 
benefits and drawbacks. Traditionally, they build on low-level R interfaces 
to graphical systems such as GTK+ [@RGtk2], Qt [@qtbase], or Java GUI frameworks 
[@rJava]. In general, the resulting system can be 
very fast and flexible, but sharing and reproducing output is usually a problem 
due to the heavy software requirements. Although there may be sacrifices in
performance, using the modern web browser as a canvas is more portable, 
accessible, and composable (graphics can be embedded within larger 
frameworks/documents).

Base R does provide a Scalable Vector Graphics (SVG) device, `svg()`,
via the Cairo graphics API [@cairo]. The R package **SVGAnnotation**
provides functionality to post-process `svg()` output in order to add
interactive and dynamic features [@SVGAnnotation]. This is a powerful
approach, since in theory it can work with any R graphic, but the
package is self-described as a proof-of-concept which
reverse-engineers poorly-structured `svg()` output. As a result, it is
not straightforward to extend this system for linked data
visualizations with advanced functionality (multiple layers, multiple
plots, multiple selection variables).

The lack of well-structured SVG for R graphics motivated the **gridSVG** package 
which provides sensible structuring of SVG output for grid graphics 
[@gridSVG]. This package also provides some low-level tools for animating 
or adding interactive features, where grid objects must be referenced by name.
As a result, use of this interface to add interactivity to a 
**ggplot2** plot requires understanding of the grid naming scheme **ggplot2** uses 
internally. An interface where 
interactivity can be expressed by referencing the data to be visualized, 
rather than the building blocks of the graphics system, would be preferable 
since the former interface is decoupled from the implementation and does not 
require knowledge of grid.

In terms of the animation API, the R package **gganimate** is very similar to our
system [@gganimate]. It directly extends **ggplot2** by adding a new aesthetic,
named `frame`, which splits the data into subsets (one for each unique 
value of the frame variable), produces a static plot for each subset, and uses
the animation package to combine the images into a key frame animation 
[@animation]. This is quite similar but not as flexible as our system's
support for animation, which we fully describe in Section \@ref(animation).
Either system has the ability to control the amount of time that a given frame 
is displayed, but our system can also animate the transition between frames via
the `d3.transition()` API [@d3]. Smooth transitions help the animation viewer track
positions between frames, which is useful in many scenarios, such as the 
touring example discussed in Section \@ref(tour). 
The **tweenr** package provides similar smooth transitions, 
by computing data values in R that interpolate between 
animation frames (in **animint**, these calculations are performed in the web browser).

The **ggvis** package is similar to our system in that is is also inspired
by the grammar of graphics [@ggvis]. It does not directly extend
**ggplot2**, but instead provides a brand new purely functional interface
which is designed with interactive graphics in mind. It currently
relies on Vega to render the SVG graphics from JSON [@vega], and the R
package **shiny** to enable many of its interactive capabilities
[@shiny]. The interface gives tremendous power to R users, as it
allows one to write R functions to handle user events. This power does
come with a cost, though, as sharing and hosting **ggvis** graphics
typically requires special web server softwares, even when the
interaction logic could be handled entirely client-side. As we outline
in Section \@ref(implementation), our system does not require a web
server, but can also be used inside **shiny** web applications, when
desired.

The tour is a useful visualization technique for exploring high-dimensional data
which requires interactive and dynamic graphics. The open-source software ggobi 
is currently the most fully-featured toolkit for touring data and has support
for interactive techniques such as linking, zooming, panning, and identifying 
[@ggobi:2007]. The R package **rggobi** [@rggobi] provides an R interface 
to ggobi's graphical interface, but unfortunately, the software requirements 
for installation and use of this toolchain are heavy and stringent. Furthermore,
sharing the interactive versions of these graphics are not possible. The R 
package **cranvas** aims to be the successor to ggobi, with support for similar
interactive techniques, but with a more flexible interface for describing 
plots inspired by the grammar of graphics [@cranvas]. **Cranvas** also has 
heavy and stringent software requirements which limits the portability and
accessibility of the software. 

Another R package for interactive graphics is **iplots** [@Urbanek2011],
which has several important differences compared to  **animint**. 
Brushing/highlighting of linked iplots is supported for
single-layer plots such as scatterplots or barplots, but it is not
easy to define new multi-layer interactive plots. Futhermore since
iplots does not use the grammar of graphics, it is difficult to create
legends and multi-panel plots. Finally since iplots requires compiled
C++ code for rendering on the local machine, its graphics are not as
easy to share as **animint** graphics which can be viewed in a web
browser.


# Extending the layered grammar of graphics

In this section, we describe in detail our extension of **ggplot2**'s layered
grammar of graphics implementation [@ggplot2-paper]. In **ggplot2**, there are five
essential components that define a layer of graphical makings: data, mappings 
(i.e., aesthetics), geometry, statistic, and position. These simple components 
are easy to understand in isolation and can be combined in many ways to express
a wide array of graphics. For a simple example, here is one way to create a
scatterplot in **ggplot2** of variables named `<X>` and `<Y>` in 
`<DATA>`:

```{r, echo = FALSE}
knitr::opts_chunk$set(eval = FALSE, message = FALSE)
```


```{r}
ggplot() + layer(
  data = <DATA>, 
  mapping = aes(x = <X>, y = <Y>), 
  geom = "point", 
  stat = "identity",
  position = "identity"
)
```

For every geometry, **ggplot2** provides a convenient wrapper around 
`layer()` which provides sensible defaults for the statistic and 
position (in this case, both are "identity"):

```{r}
ggplot() + geom_point(
  data = <DATA>, 
  aes(x = <X>, y = <Y>)
)
```

A single **ggplot2** plot can be comprised of multiple layers, and different layers 
can correspond to different data. Since each graphical mark within a **ggplot2** 
layer corresponds to one (or more) observations in `<DATA>`, aesthetic
mappings provide a mechanism for mapping graphical selections to the original 
data (and vice-versa) which is essential to any interactive graphics system [@plumbing]. 
Thus, given a way to combine multiple 
**ggplot2** plots into a single view, this design can be extended to support a 
notion of multiple linked views, as those discussed in @Ahlberg:1991 
and @Buja:1991vh.

## Linking views via aesthetic mappings {#extension}

<!--
Direct manipulation, as discussed in @Ahlberg:1991, is a graphical
interface for interacting with databases. Direct manipulation
interfaces include a graphical representation of queries to the
database, which can be manipulated using the mouse. In the context of
statistical graphics, direct manipulation refers to interfaces in
which clicking the plotted representation of the data (such as lines
or points) changes the plot. In contrast, indirect manipulation
interfaces use widgets (such as buttons or menus) to change the plot.
-->

@ggobi:2007 use `SQL` queries to formalize the linked views infrastructure
discussed in @Ahlberg:1991 and @Buja:1991vh. We use a similar approach 
to show how aesthetic mappings can be used to assign data values to graphical
marks via **ggplot2** to support similar graphical queries. It's worth noting 
that, since these aesthetics effectively define a set of database queries
that are known at print time, these queries can be made by directly 
manipulating graphical marks and/or indirectly via a dropdown widget, 
as discussed in Section \@ref(direction). It's also worth noting that these 
aesthetics could be defined in such a way that they are not solely restricted 
to any particular direct manipulation event (e.g. mouse click), but for sake of demonstration,
we restrict focus to our **animint** implementation, which has `clickSelects` 
and `showSelected` aesthetics. 

As demonstrated in Figure \@ref(fig:tips), the `R` code used to generate it,
and the corresponding `SQL` queries, `clickSelects` assigns data to graphical 
marks, which when clicked, sets selection value(s).\footnote{Interactive versions
of all of the figures mentioned in this paper can be found at
\url{http://members.cbio.mines-paristech.fr/~thocking/animint-paper-figures/}} 
For this example, we refer to the current set of selection value(s) (i.e., 
the selection set) as `selected_sex`. The value of `selected_sex` can take on
any subset of the unique values of `sex`, and in this case, the unique values 
are `Male` and `Female`. The selection set, `selected_sex`, is then matched 
against graphical marks containing a `showSelected` property. As a result,
by clicking on a bar in Figure \@ref(fig:tips), one may highlight the 
relationship between `tip` and `total_bill` for either gender.

```{r}
library(animint)
data(tips, package = "reshape2")
bar <- ggplot(tips) + 
  geom_bar(aes(x = sex, clickSelects = sex))
scatter <- ggplot(tips) +
  geom_point(aes(x = total_bill, y = tip), alpha = 0.3) +
  geom_point(aes(x = total_bill, y = tip, showSelected = sex))
animint2dir(list(bar = bar, scatter = scatter))
```

```{r tips, echo = FALSE, eval = TRUE, fig.cap = "A graphical query of tips data set. Left: the `clickSelects` aesthetic designates a clickable geom bar that can change a selection variable. Right: the `showSelected` aesthetic designates a geom point that responds by showing only the data which corresponds to the current selection."}
knitr::include_graphics("images/tips")
```

In the R code above, the `geom_bar()` layer in the left-hand plot is linked to 
the 2nd `geom_point()` layer in the right-hand plot since the `clickSelects` and
`showSelected` aesthetics are mapped to a common variable, `sex`. Note how the 
first `geom_point()` layer does not have a `showSelected` mapping, but has
a bit of alpha transparency, so all the data is shown in light gray, and the
current selection is portrayed in black. In other words, when a bar is clicked, 
in order to update the second layer of points, our system performs an SQL query 
of the form:

```sql
SELECT * FROM tips
  WHERE sex IN selected_sex
```

In this example, `selected_sex` is either `Male` or `Female` (a single selected value), 
but as we show in later examples, a selection set can also can multiple values.
Although the `clickSelects` aesthetic is tied to a mouse click event,
other aesthetics could easily be created to support other selection
events, such as hover or click+drag. Statistically speaking, this type
of interaction is useful for navigating through joint distributions
conditional upon discrete values. In this sense, our extension is
closely related to trellis displays [@trellis] and linked scatterplot
brushing [@brushing-scatterplots]. The major differences are that our
conditioning is layer-specific (not plot-specific), is not tied to a
particular geometry, and can be controlled through direct manipulation
or animation controls.

<!-- 
TODO: maybe touch on the fact that we can't dynamically recompute statistical 
aggregations, but you can always pre-compute?
-->

## World Bank example {#worldbank}

<!--
%% In the system we propose, the central concept of interactivity is a
%% selection variable, such as year or region in Figure~1. For each
%% selection variable, one or several values can be selected at a time,
%% e.g. year=1979 and country=\{United States, Vietnam\}.  Like
%% Cleveland's system, we also use enhanced linking to highlight the
%% selected value(s) of each selection variable. In contrast to
%% Cleveland's single rectangular brush that selects points in plots of a
%% single data table, we propose several selection variables in linked
%% plots of several data tables. Linking is accomplished using common
%% names when declaring \texttt{clickSelects} and \texttt{showSelected}
%% selection variables.

%% For example, to declare a clickable plot element that changes the
%% selected value of the year variable, a designer writes
%% \texttt{clickSelects=year} (tallrect in Figure~1). And to show only
%% the data subset for the selected value of year, a designer writes
%% \texttt{showSelected=year} (point and text in Figure~1).

%% Using just the \texttt{clickSelects} and \texttt{showSelected}
%% keywords, a wide variety of interactive visualizations can be
%% defined. To make the selection automatically change over time
%% (animation), a designer may declare one variable as the \texttt{time}
%% variable. Since it is perceptually advantageous to have smooth
%% transitions in data-driven animations \citep{animated-transitions}, a
%% designer may also declare a \texttt{duration} list of selection
%% variables which should have smooth transitions.
-->

This section uses the linking framework introduced in the previous section
to a more complex data set provided by the World Bank. 
The interactive version of Figure \@ref(fig:worldbank) fosters
exploration of the relationship between life expectancy and fertility
over time for 205 countries. The year 1979 and the
countries United States and Vietnam are selected in the static version of 
Figure \@ref(fig:worldbank), but readers are
encouraged to change the selection by clicking on the interactive
version, which is provided in the supplementary materials. The
interactive version also makes use of additional options
to automatically increment the selected year every few seconds, allowing
us to visualize the evolution of the relationship between life expectancy
and fertility rate through animation. These options are explained in detail in 
Section \@ref(animation).

```{r worldbank, echo = FALSE, eval = TRUE, fig.cap = "An interactive animation of World Bank demographic data of several countries, designed using `clickSelects` and `showSelected` keywords (top).  Left: a multiple time series from 1960 to 2010 of life expectancy, with bold lines showing the selected countries and a vertical grey tallrect showing the selected year. Right: a scatterplot of life expectancy versus fertility rate of all countries. The legend and text elements show the current selection: year=1979, country={United States, Vietnam}, and region={East Asia & Pacific, North America}"}
knitr::include_graphics("images/figure-1")
```

We anticipate that some **ggplot2** users will be able to reverse engineer the
code which creates Figure \@ref(fig:worldbank), simply by looking at it.
In fact, this is a big reason why **ggplot2** is so widely used: 
it helps minimize the amount of time required to translate an idea for a figure into computer code. 
Note that, in the left hand plot of 
Figure \@ref(fig:worldbank), we have a time series 
of the life expectancy where each line is a country (i.e., we `group` by
country) and lines are colored by region. By clicking on a line, we also want 
the country label to appear in the right hand plot, so we also need to set 
`clickSelects=country`. Lastly, by setting `showSelected=region` and 
`color=region`, we can hide/show lines by clicking on the color legend entries.

```{r}
timeSeries <- ggplot() + geom_line(
  data = WorldBank,
  aes(x = year, y = life.expectancy,
      group = country, color = region,
      clickSelects = country, 
      showSelected = region)
)
```

To help point out the currently selected year, we also provide a visual cue in 
the form of tall rectangles to the time series plot.
These tall rectangles will also serve as a way to directly modify the
selected year. The tallrect geometry is a special case of a rectangle
that automatically spans the entire vertical range, so we just have to
specify the horizontal range via `xmin` and `xmax` aesthetics. Also, since the
layered grammar of graphics allows for different data in each layer,
we supply a data frame with just the unique years in the entire data
for this layer.

```{r}
years <- data.frame(year = unique(WorldBank$year))
timeSeries <- timeSeries + geom_tallrect(
  data = years,
  aes(xmin = year - 0.5, xmax = year + 0.5,
      clickSelects = year)
)
```

As for the right hand plot in Figure \@ref(fig:worldbank), there are 
three layers: a point layer for countries, a text layer for countries, 
and a text layer to display the selected year. By clicking on a point,
we want to display the country text label and highlight the corresponding
time series on the left hand plot, so we set `clickSelects=country`
in this layer. Furthermore, we only want to show the points for the 
selected year and region, so we also need `showSelected=year` and 
`showSelected2=region`. 

```{r}
scatterPlot <- ggplot() + geom_point(
  data = WorldBank,
  aes(x = fertility.rate, y = life.expectancy,
      color = region, size = population,
      clickSelects = country,
      showSelected = year,
      showSelected2 = region)
)
```

Note that any aesthetics containing the substring `showSelected`
(including `showSelected2`) are interpreted as `showSelected`
variables, and combined together using the intersection operation. In
the example above, that means that a point will be drawn for the
currently selected combination of year and region, as in the following SQL query,

```sql
SELECT * FROM WorldBank
  WHERE year   IN year_selection
  AND   region IN region_selection
```

Below, the text layer for
annotating selected countries is essentially the same as the point
layer, except we assign the country name to the `label` aesthetic.

```{r}
scatterPlot <- scatterPlot + geom_text(
  data = WorldBank,
  aes(x = fertility.rate, y = life.expectancy,
      label = country,
      showSelected = country,
      showSelected2 = year,
      showSelected3 = region)
)
```

Lastly, to help identify the selected year when viewing the scatterplot, 
we add another layer of text at a fixed location.

```{r}
scatterPlot <- scatterPlot + geom_text(
  data = years, x = 5, y = 80,
  aes(label = paste("year =", year),
      showSelected = year)
)
```

In summary, this section shows an example of how the proposed
`clickSelects` and `showSelected` aesthetics can be used with several
different geoms (line, point, text, tallrect), each of which can potentially
display a different data set. In each case we use `clickSelects` to
declare a geom that when clicked updates the current selection, 
and we use `showSelected` to declare a geom which responds
to such changes by updating the set of displayed data. In the next
sections, we further options that allow us to accumulate selections,
update selections indirectly, and automate selection updates.

## Linking and multiple selection {#linking}

Linking is declared in R code by putting ggplots with common `clickSelects` and
`showSelected` aesthetics together in a list. For example, we can link
the ggplots from the previous section by including them together in the following list:

```{r}
viz <- list(
  timeSeries = timeSeries,
  scatterPlot = scatterPlot
)
```

Linking is accomplished because the two ggplots declared
`clickSelects` and `showSelected` aesthetics that refer to common
variable names (`region`, `year`, `country`). For each such selection
variable, our system updates the set of selected values
in response to mouse clicks on `clickSelects` geoms, and then updates
the corresponding data which is displayed for `showSelected`
geoms.

Note that the `viz` list above can also contain numerous
options which are listed in Table \@ref(tab:overview). For example, the
`selector.types` option controls whether or not selections for a given
variable accumulate (single or multiple selected values). This sort of logic
has also been interpreted as transient versus persistent selection [@ggobi:2007].

```{r}
viz$selector.types <- list(
  year = "single",
  country = "multiple",
  region = "multiple"
)
```

The code above declares `year` as a single selection
variable, which means that only a single year may be selected at a
time (clicking a geom with `clickSelects=year` will change the
selection to the corresponding year). The `country` and `region`
variables are declared as multiple selection variables, which can have
multiple selected values at a time (clicking a geom with
`clickSelects=country` will add/remove that country to/from the
selection set). 

## Direct versus indirect manipulation {#direction}

When viewing the interactive version of Figure \@ref(fig:worldbank), suppose
we wish to select Thailand. Direct manipulation is not very useful in 
this case since it is not easy to identify and select Thailand based
on graphical marks on a plot. For this reason, **animint** also provides 
dropdown menu(s) for each selection variable to aid the selection process.
Figure \@ref(fig:widgets) shows what the user sees after typing ''th'' in 
the search box. Note that these dropdowns support selection of multiple
values and coordinate sensibly with selections made via direct manipulation.

```{r widgets, echo = FALSE, eval = TRUE, fig.cap = "Animint provides a menu to update each selection variable. In this example, after typing \"th\" the country menu shows the subset of matching countries."}
knitr::include_graphics("images/dropdown")
```

## Animation and smooth transitions  {#animation}

Animation is declared using the `time` option, which specifies a
selection variable which will be automatically updated over time, as
well as a time delay in milliseconds. The code below declares the
`year` variable to be animated every 3 seconds. 

```{r}
viz$time <- list(variable = "year", ms = 3000)
```

Animation is useful in the World Bank data visualization because it shows how
the bi-variate relationship between fertility rate and life expectancy
changes over time. Animation clearly shows how many countries progress
from low life expectancy and high fertility rate in early years, to
high life expectancy and low fertility rate in later years.

Finally, the `duration` option specifies the amount of time used to
smoothly transition between selections (with linear easing). Smooth
transitions help the viewer track geoms before and after an update to
the selection set. For example in the code below we declare a 1 second
smooth transition on the `year` variable, in order to more easily
track the points on the scatterplot. 

```{r}
viz$duration <- list(year = 1000)
```

Note that for accurate interpretation of smooth transitions, the new
`key` aesthetic must be specified. The `key` aesthetic is used to
match data elements before and after the smooth transition. In the
World Bank example, we would need to specify `aes(key=country)` for
the points and text in the scatterplot.

## Storing and re-storing state

When sharing an interactive visualization with others, it can often be helpful
to share interesting state(s) of the visualization. In **animint**, states can
be serialized in a URL link and/or specified at the command line via the `first`
option. The code below declares that the first selection of the `country` 
variable is the set of two countries, United States and Vietnam.

```{r}
viz$first <- list(country = c("United States", "Vietnam"))
```

## Compiling and rendering

Supplying the `viz` list of ggplots and rendering options to
the `animint2dir()` function will save all the files necessary for
rendering the visualization:

```{r}
animint::animint2dir(viz)
```

As shown in Supplementary Figure 1, the **animint** system consists of 
2 parts: the compiler and the renderer. The compiler is R code that 
converts a list of ggplots and options to a JSON plot meta-data file 
and a tab-separated values (TSV) file database. The renderer consists 
of HTML and JavaScript files, which
can be easily hosted along with the TSV and JSON files on any web
server. The interactive plots can be viewed by opening the
`index.html` page in any modern web browser. Note that **animint** 
currently depends on a fork of
**ggplot2**\footnote{\url{https://github.com/faizan-khan-iit/ggplot2/tree/validate-params}}
that contains some minor modifications which are needed to support
interactive rendering on web pages. Additional implementation details
are available in the supplementary materials.

# Exploring scope with examples {#performance}

This section attempts to demonstrate the range of visualizations that
are supported by the system we propose. In particular because of its
support for interaction and animation, it excels at display of
interactive maps with time-varying data. We give two such examples
below.

## Tornadoes in the United States 

One of the strong points of the system we propose is display of
multi-layer plots such as maps with time-varying data. For example,
Figure \@ref(fig:tornado) shows a visualization of US tornado data
from 1950 to 2012. This data visualization consists of two multi-layer
plots with two interaction variables, `year`, and `state`. 

The left plot is a map which shows state borders using a polygon with
`clickSelects=state`. The currently selected state is shown using
semi-transparency, and other states can be selected by clicking
them. The state map plot uses geoms with `showSelected=year` to show
tornado paths (segment geom) and endpoints (point geom) for the
currently selected year (which is emphasized with a text geom above the map). 

The right plot uses several geoms to show details for the currently
selected state and year. A bar geom shows a time series of tornado
counts for the selected state (`showSelected=state`), which can be
clicked to change the currently selected year (`clickSelects=year`). A
text geom at the top of the plot shows the currently selected state
(`showSelected=state`), and a text geom at the bottom emphasizes the
tornado count for the selected year (using `showSelected` variables
for both `state` and `year`).

These interactions can be useful for discovering patterns in
the data, and for suggesting models that can describe or predict tornado
paths.


```{r tornado, echo = FALSE, eval = TRUE, fig.cap = "Interactive animation of US tornadoes from 1950 to 2012. This diagram depicts a scenario where the user queried Texas (by clicking the map), and the year 1982 (by clicking the bar chart).  In addition to the graphical elements being highlighted as a visual clue of what query is being made, this visualization includes dynamic text labels reflecting the query."}
knitr::include_graphics("images/figure-tornado")
```

## Central American climate data

```{r climate, echo = FALSE, eval = TRUE, fig.cap = "Visualization containing 6 linked, interactive, animated plots of Central American climate data. Top: for the selected time (December 1997), maps displaying the spatial distribution of two temperature variables, and a scatterplot of these two variables. The selected region is displayed with a black outline, and can be changed by clicking a rect on the map or a point on the scatterplot. Bottom: time series of the two temperature variables with the selected region shown in violet, and a scatterplot of all times for that region. The selected time can be changed by clicking a background tallrect on a time series or a point on the scatterplot. The selected region can be changed by clicking a line on a time series."}
knitr::include_graphics("images/figure-climate")
```

A more complex map data visualization example is shown in Figure
\@ref(fig:climate), which depicts climate time series data observed in
Central America. There are two interaction variables, `time` and `region`. 

Two maps in the upper right display borders of
the countries in and near Central America. Unlike the previous
example with US states, the country borders are static (clicking has no effect). For the currently
selected time, rect geoms with `showSelected=time` show the spatial distribution of sea surface temperature as
well as its deviation from the monthly norm. Since `clickSelects=region` is specified, 
clicking a rect changes the currently selected region, 
which is emphasized with a black border. 
These plots facilitate visualization of the spatial distribution of the climate variables,
and how they change over time.n

The plots below the maps use lines to show time series of the climate variables. 
Since `clickSelects=region` is specified, 
clicking a line changes the currently selected region, 
which is emphasized with a purple color. 
A semi-transparent tallrect shows the currently selected time;
other tallrects can be clicked to update the time (`clickSelects=time`).
These plots make it easy to select different times and regions, 
and to make comparisons between times and regions.

Scatterplots on the right use `showSelected` variables with point and text geoms,
to show the joint distribution of the two
temperature variables for the selected time (top) and region (bottom). 
The plots use `clickSelects` to emphasize the currently selected region (top) and time (bottom), 
and are useful for visualizing normality and outliers in the joint distribution.

# Limitations and future work {#limitations}

The system we have proposed enables graphical queries and animation via an 
extension to **ggplot2** using entirely client-side web technologies. Often times
in interactive statistical graphics, we wish to dynamically compute a statistical 
aggregation(s) as a function of a user selection (e.g., compute a new 
linear model based on a set of newly brushed points). Using such a client-side 
linking framework requires one to pre-compute every 
possible aggregation ahead of time. If the number of states is fairly small, this isn't 
much of a problem, but it doesn't scale very well. For an example of pre-computing
states and exploring those limitations, see the touring example in the supplementary 
materials.

It's possible that future implementations could dynamically recompute
statistical aggregated specified via a **ggplot2** layer, but it would be an 
immense amount of work to translate so many `R` functions to `JavaScript`. 
At this point,


Our current implementation provides a visual indication of the current selection
via semi-transparency of `clickSelects` geoms. In future work we would
like to explore more obvious visual cues that can be used to quickly
show the user the links between plots and possible interactions.

A number of limitations in our current implementation derive from the
fact that some plot features are computed once during the compilation
step, and remain static on a rendered plot. For example, users are
unable to change variable mappings after compilation.  Also, when
different data subsets have very different ranges of values, it may be
preferable to recompute scales when `clickSelects` selection(s)
change. Some of these limitations can be resolved by adding
interactive widgets to "recompile" components hard-coded in the plot
meta information. In fact, **animint** makes it easy to embed
visualizations inside of **shiny** web applications, and we have an
example of interactively redefining variable mappings.

Our compiler also currently takes advantage of **ggplot2** internals to compute 
statistics and positional adjustments before rendering. As a result,
statistics/positions will not dynamically recompute based on selections.
In other words, using `clickSelects`/`showSelected` with 
non-identity statistic(s)/position(s) may not generate a sensible result. 
It would be possible, but a significant amount of work, to transfer these 
computations from the compiler to the renderer.

Another set of limitations derive from our current restriction that all 
subsets (corresponding to each possible selection) must be precomputed
before render time. As elucidated in Section \@ref(tour),
if there is a large space of possible selections,
it is impractical to precompute every subset before viewing.
Therefore, for future work it would be useful if the renderer could
dynamically compute subsets when new selections are made. 

Our implementation is also limited to two specific types of
direct manipulation: selecting graphical elements via
mouse click (`clickSelects`), and showing/hiding related 
elements (`showSelected`). However, the framework described in 
Section \@ref(extension) is not restricted to a particular event type, 
so `hoverSelects` and `brushSelects` aesthetics could be 
added, for instance. There are other types of interaction that could be
added, that wouldn't require additional extensions to the grammar of graphics,
such as: zooming, panning, and plot resizing.

# Conclusion

We have proposed several extensions to **ggplot2**'s layered grammar of graphics 
in order to support a declarative approach to producing interactive
and dynamic web graphics. By adding clickSelects and showSelected aesthetics to specify selection
source(s) and target(s), **ggplot2** users can quickly and easily create 
animations with smooth transitions and perform dynamic queries via direct
manipulation of linked views. As a result, **animint** is a useful tool 
not only for EDA, but also for the presentation
and distribution of interactive statistical graphics.

\section*{Interactive figures and reproducible research statement}

The source code to create this paper and its figures is online at
https://github.com/tdhock/animint-paper/ and the interactive figures
can be viewed at
http://members.cbio.mines-paristech.fr/~thocking/animint-paper-figures/

\section*{Acknowledgements}

The authors wish to thank **animint** users MC Du Plessis, Song Liu,
Nikoleta Juretic, and Eric Audemard
who have contributed constructive criticism and helped its development.

\section*{References}
